---
title: "The riverine fauna" 
author:  "Angela Cukusic"

output:
  html_document:
    toc: true               # Table of contents
    toc_float: true         # Keep the table of contents floating on the page
    code_download: true     # Enable code download button
    code_copy: true         # Enable code copy button
    number_sections: true   # Numbering the sections
    theme: cerulean           # Choose a Bootstrap theme (check available themes)
    highlight: tango        # Syntax highlighting style 
---

```{r setup, include=FALSE}
# For width of code chunks and scroll bar 
options(width=250)

knitr::opts_chunk$set(eval = TRUE, 
                      echo = TRUE, 
                      cache = TRUE,
                      include = TRUE,
                      warning = FALSE,
                      collapse = FALSE,
                      message = FALSE,
                      engine = "R", 
                      # Chunks will always have R code, unless noted
                      error = TRUE)

```


# load packages
```{r}
#if some are not already installed you will gwt an error when loading them with library()
#in that case use nstall.packages("name.of.package")
library(readxl) # to read in datasets that are in xlxs form
library(tidyverse) # to edit datasets easily
library(dplyr)  # to edit datasets easily
library(vegan)  # to use ecological analyses
library(patchwork) # to show more plots in the same picture

```


```{r}
# palette
c27 <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "violet", # orange
  "black", "gold1",
  "darkturquoise","#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", 
  "deeppink1",# lt purple
  "#FDBF6F", # lt orange
  "gray70", "khaki2",
  "maroon","yellow4", "orchid1",  "blue1", 
 "skyblue2", "green1",  "yellow3",
  "darkorange4","steelblue4",  "brown",
  "blue4", "#FF7F00"
)
```


# load data
```{r}

env_vj <- read_excel("C:/Users/Angela Cukusic/Documents/faks/joanna/Environment_Vjosa_v2.xlsx") %>% 
          as.data.frame() %>%  # to turn it into the form that we can work with
          column_to_rownames("Code") # assign rownames 

spec_vj <- read_excel("C:/Users/Angela Cukusic/Documents/faks/joanna/Species_Vjosa_v1.xlsx") %>% 
           as.data.frame() %>%
           column_to_rownames("Code") %>% 
           mutate_at(-c(1,2), as.numeric) # species dataset does not think of 1-2-3 as numbers but letters, make it numeric
  # "%>%" and mutate are all commands from tidyverse, see https://www.r-bloggers.com/2020/06/a-very-short-introduction-to-tidyverse/ 

spt_vj <- read_excel("C:/Users/Angela Cukusic/Documents/faks/joanna/Spatial_vjosa_v3.xlsx") %>% 
          as.data.frame() %>% 
          column_to_rownames("Code") %>% 
          mutate_at(-c(1,2,3), as.numeric) # to make coordinates numeric



# for later
spec_vj_no_na <- spec_vj[complete.cases(spec_vj) , -c(1,2)] #remove nas

# have all three datasets combined, and that without the NAs
all_data <-  left_join(rownames_to_column(spec_vj_no_na), 
                       rownames_to_column(env_vj), 
                       by = "rowname") %>% 
             left_join(. , rownames_to_column(spt_vj), 
                       by = "rowname" ) %>% 
             column_to_rownames("rowname") %>% 
  # selecting with "!" means selecting things that are NOT named 
             select(! c("SampleID.y", "River.y", "Month.y")) %>% 
             rename("SampleID" ="SampleID.x", "River" = "River.x", "Month" = "Month.x")
```


# alpha diversity

## calculate 
```{r}

# add the site identification
all_data <- all_data %>% mutate(Site = case_when(str_detect(SampleID, "GWA") ~ "A",
                                                  str_detect(SampleID,"GWB") ~ "B"))
#with this we create another column where it says whether something is A or B site
# with mutate you add a new column

#pg 16
#calculating richness and shannons index
richness <- specnumber(spec_vj_no_na)
shannondiv <- diversity(spec_vj_no_na)
effective_number_species <- exp(shannondiv)

```


## test
```{r}

# 1. richness

#test with anova
summary(aov(richness ~ all_data$River)) #not significantly different
summary(aov(richness ~ all_data$Site)) #not significantly different

#KW test
kruskal.test(richness ~ all_data$River) #not significantly different
kruskal.test(richness ~ all_data$Site) #not significantly different

# which test to use: anova or KW
#if shapiro.test > 0.05 and bartlett.test > 0.05 use anova, if not use KW
shapiro.test(richness) 
bartlett.test(richness ~ River, data = all_data)
#...check for ~ Site and for shannons index and so on ... 


#important: although Langarica and Kardhiq seem different it is important to remember that:
table(all_data$River)
#.. Langarica has only 2 data points (after we took out NAs) which could impact the results




# 2. shannons index

#seeing if it depends on the environmental variables
boxplot(shannondiv ~ all_data$River)
boxplot(shannondiv ~ all_data$Site)

#test with anova
summary(aov(shannondiv ~ all_data$River)) 
summary(aov(shannondiv ~ all_data$Site)) 

#KW test 
kruskal.test(shannondiv ~ all_data$River) 
kruskal.test(shannondiv ~ all_data$Site) 

#if shapiro.test > 0.05 and bartlett.test > 0.05 use anova, if not use KW
shapiro.test(shannondiv) # is not normal - maybe we could do better to test with KW
bartlett.test(shannondiv ~ River, data = all_data)
 

```

## plot

```{r}

by_river <- cbind(as.data.frame(richness), as.factor(all_data$River)) %>% 
  as.data.frame() %>%
  rename("div"= "richness","river"="as.factor(all_data$River)") %>%
  ggplot( aes(x = river, y = div, fill = river, color=river)) +
  geom_jitter(width = 0.2, size = 3, 
              shape=21, alpha = 0.3,
              aes(fill = river) , 
              color = "black") +
  geom_boxplot( alpha=0.5, outlier.shape = NA, 
                aes(fill = river)) + 
  theme_minimal()+
  theme(legend.position = "none",
        axis.title=element_text(size=8,face="bold"),
        #axis.text.x = element_text(angle = 90)
        axis.text.y = element_text(size=8,face="bold")) +
  labs(x = "River",
       y = "Richness") 

by_site <- cbind(as.data.frame(richness), as.factor(all_data$Site)) %>% 
  as.data.frame() %>%
  rename("div"= "richness","site"="as.factor(all_data$Site)") %>%
  ggplot( aes(x = site, y = div, fill = site, color=site)) +
  geom_jitter(width = 0.2, size = 3, 
              shape=21, alpha = 0.3,
              aes(fill = site) , 
              color = "black") +
  geom_boxplot( alpha=0.5, outlier.shape = NA, 
                aes(fill = site)) + 
  theme_minimal()+
  theme(legend.position = "none",
        axis.title=element_text(size=8,face="bold"),
        #axis.text.x = element_text(angle = 90)
        axis.text.y = element_text(size=8,face="bold")) +
  labs(x = "Site",
       y = "Richness")  + 
  scale_x_discrete(limits = c("A", "B") ) #remove NA


by_river_s <- cbind(as.data.frame(shannondiv), as.factor(all_data$River)) %>% 
  as.data.frame() %>%
  rename("div"= "shannondiv","river"="as.factor(all_data$River)") %>%
  ggplot( aes(x = river, y = div, fill = river, color=river)) +
  geom_jitter(width = 0.2, size = 3, 
              shape=21, alpha = 0.3,
              aes(fill = river) , 
              color = "black") +
  geom_boxplot( alpha=0.5, outlier.shape = NA, 
                aes(fill = river)) + 
  theme_minimal()+
  theme(legend.position = "none",
        axis.title=element_text(size=8,face="bold"),
        #axis.text.x = element_text(angle = 90)
        axis.text.y = element_text(size=8,face="bold")) +
  labs(x = "River",
       y = "Shannon index") 

by_site_s <- cbind(as.data.frame(shannondiv), as.factor(all_data$Site)) %>% 
  as.data.frame() %>%
  rename("div"= "shannondiv","site"="as.factor(all_data$Site)") %>%
  ggplot( aes(x = site, y = div, fill = site, color=site)) +
  
  geom_jitter(width = 0.2, size = 3, 
              shape=21, alpha = 0.3,
              aes(fill = site) , 
              color = "black") +
  geom_boxplot( alpha=0.5, outlier.shape = NA, 
                aes(fill = site)) + 
  theme_minimal()+
  theme(legend.position = "none",
        axis.title=element_text(size=8,face="bold"),
        #axis.text.x = element_text(angle = 90)
        axis.text.y = element_text(size=8,face="bold")) +
  labs(x = "Site",
       y = "Shannon index") + 
  scale_x_discrete(limits = c("A", "B") ) #remove NA


by_river +  by_site + by_river_s  + by_site_s + plot_layout(widths = c(2, 1))

```


# beta diversity (community data)

## calculate
```{r}
# do an ordination with a hellinger transformation 
 sp.dist <-  vegdist(decostand( spec_vj_no_na,
                               method = "hellinger") )


# or without the transformation
sp.dist <-  vegdist(
  spec_vj_no_na
  )

nmds <- metaMDS(sp.dist, k = 2, autotransform = FALSE)
```

## plot
```{r}
nmds_plot <- as.data.frame(nmds$points) %>% 
  rownames_to_column("Locality") %>%
  left_join(., spt_vj %>%   rownames_to_column("Locality"), 
            by = "Locality") %>% 
  mutate(River = as.factor(River)) %>% 
  ggplot(aes(x=MDS1,y=MDS2) ) +
  geom_point(aes( color= River, shape = Month),
             size = 4, alpha = 0.6, stroke = 0.5) +  

  theme_bw() +
# add stress value
  annotate(geom="text",x=0.6, y=2, 
           label="stress =", color="black") +   
  annotate(geom="text", x=1.1, y=2, 
           label=round(nmds$stress,4), color="black") +
# ellipse  
  stat_ellipse(aes(linetype = Month), linewidth = 0.4) + 
# visual parameters 
  theme(legend.background = element_blank(),
        legend.box.background = element_rect(colour = "gray")) +
  labs(fill = "Temperature \ncategory") +
  geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
  geom_hline(yintercept = c(0), color = "grey70", linetype = 2) 

nmds_plot

```



## test
```{r}
# a) for river

# 1. test dispersion as an assumption 
# (variance test for multivariate statistics)
spec_vj_no_na %>% rownames_to_column("Locality") %>% 
    left_join(., spt_vj %>%   rownames_to_column("Locality"), 
            by = "Locality") %>% 
  dplyr::select("River", "Month") -> no_na_categorical
disp <- betadisper( sp.dist, no_na_categorical$River )  



# p>0.05 is what we need - homogenous variances
anova(disp) # no differences in dispersion 


# 2. now test actual ordination
# test differences in group centroid with adonis 
# (anova version for multivariate statistics)
permanova <- adonis2(sp.dist ~ no_na_categorical$River, 
                     na.action = na.omit,
                     method = "bray" , permutations = 999)
permanova # not sign.


# b) for month

# 1. test dispersion as an assumption 
# (variance test for multivariate statistics)
disp <- betadisper( sp.dist, no_na_categorical$Month )  


# p>0.05 is what we need - homogenous variances
anova(disp) # no differences in dispersion 


# 2. now test actual ordination
# test differences in group centroid with adonis 
# (anova version for multivariate statistics)
permanova <- adonis2(sp.dist ~ no_na_categorical$Month, 
                     na.action = na.omit,
                     method = "bray" , permutations = 999)
permanova # sign. differences between the two sampling campaigns

```


# environmental data

### rda
```{r}

# make clean environmental data
env_vj %>% dplyr::select(!c("SampleID", "Date", "Month", "River")) %>% 
  rownames_to_column("ID") -> env1

  
spt_vj  %>% rownames_to_column("ID") %>% 
  dplyr::select(!"SampleID") -> env2

full_join(env1, env2, by = "ID") -> env
# choose only  the ones that we used for species matrix
env[which(env$ID %in% rownames(spec_vj_no_na) ),] %>% 
  select(!c("River", "Month", "ID", "Air-Temp_C") ) %>% 
  mutate_if(is.character, as.numeric ) %>% 
  scale() %>% 
  as.data.frame() -> env.2        

# 1. first model
simpleRDA <- rda(decostand( spec_vj_no_na,
                               method = "hellinger")  ~  ., data = env.2 ,
                 distance = "bray",
                 na.action = na.omit)
# take only that are vif<10
env_vars_low_vif <- vif.cca(simpleRDA) %>% 
                    as.data.frame() %>% 
                    dplyr::rename("env" = ".") %>% 
                    dplyr::filter(env < 10) 

# make new environmental data with only low vif variables
env.2[,which(colnames(env.2) %in%  rownames(env_vars_low_vif) )] -> env.dat


# 2. model without them
simpleRDA <- rda(decostand( spec_vj_no_na,
                               method = "hellinger")  ~  ., 
                      data=  env.dat ,
                 distance = "bray",
                 na.action = na.omit) 


# test all variables and axes
# canonical axes
anova.model <- anova.cca(simpleRDA, by='axis', step=1000)  
# no axes sign

# the whole model
anova.model2 <- anova.cca(simpleRDA, step=1000) #is sign
RsquareAdj(simpleRDA)$adj.r.squared # and explains <1%

# env parameters
anova.model3 <- anova(simpleRDA, step=1000, by = "term") 
# none sign


plot(simpleRDA)
scores(simpleRDA)$species
scores(simpleRDA)$sites
scores(simpleRDA)$biplot
```


## plot
```{r}

# use the automatic plot but then you cant color by group  
  ggvegan:::autoplot.rda(simpleRDA) 

  
  # vectors
ccavectors <- as.matrix(scores(simpleRDA, display = "bp", scaling = "sites") * 3) %>% 
  t() %>% as.data.frame() %>% 
  rename( "Nitrat" = "Nitrat_mgL", 
         "NO2" = "N_NO2_ugL" ,
         "NH4" = "N_NH4_ugL",
         "pH" = "ph" ,
          "oxygen" = "O2_saturation",
         "Temp" = "GW_Temp_C", 
         "Altitude" = "Altitude_m" ) %>%
  t() %>% as.data.frame()


# site coordinates
site_data <- scores(simpleRDA, display = "sites") %>% 
  as.data.frame() %>% 
  rownames_to_column("ID") %>% 
  left_join(.,   all_data %>%    rownames_to_column("ID") %>% select("River", "Month", "ID"),
            by = "ID")

# plotting
plot_cca <- 
  site_data %>% 
  ggplot( aes(x = RDA1, y = RDA2)) +
  geom_point(aes(color = River, shape = Month), 
             size = 5, alpha= 0.5)+  
  geom_segment(data = ccavectors, aes(x = 0, y = 0, xend = RDA1, yend = RDA2), 
               size = 1.1, arrow = arrow(length = unit(0.5, "cm"))) +
   geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
   geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +
  geom_text(data = ccavectors, aes(x = RDA1*1.19, y = RDA2*1.19, 
                                   label = rownames(ccavectors)),
                                   size=6, color= "grey7" ) +
  theme_bw() +
     theme(panel.grid = element_blank(),
       legend.background = element_blank(),
           legend.box.background = element_rect(colour = "gray")) 

plot_cca
  
  

```
Dishinica has elevated levels of Nitrat.


# composition
```{r}
# detach(package:plyr, unload = TRUE)
# tried fixing conflicts from select and filter by detaching
# but it is a conflict with some other package as well
# simply continue adding dplyr:: before the command 

all_data[,1:30]    %>%
  
  select(!c("SampleID", "Date")) %>% 
  mutate(River = all_data$River) %>% 
  pivot_longer(cols=-c("Month", "River"), 
               values_to = "abundance", 
               names_to = "species") %>%
    
  
# to find relative abundance within each river  
  dplyr::group_by(Month,River) %>%
  mutate (rel_abund = abundance / sum(abundance)) %>%
  ungroup() %>%
 
# plot
  ggplot(aes(x = River, y = abundance, fill = species)) + 
  geom_bar(stat = "identity") +
  facet_wrap(~Month, scales= "free") +
  theme_bw() +
  scale_fill_manual(values=c27) +
  theme(panel.grid = element_blank(), legend.position="right",
        legend.text = element_text(size = 6),      # Adjust the text size
        legend.title = element_text(size = 6),
        legend.key.size = unit(0.4, "cm"),
        axis.title=element_text(size=8),
        axis.text.y = element_text(size=8),
        axis.text.x = element_text(angle = 90)) + 
  labs(y = "Relative abundance", 
       fill = "Class") +
  guides(fill=guide_legend(ncol = 2))
```



# along the river

```{r}
# first plot the river alone
```


For species most abundant:
```{r}
all_data %>% 
  ggplot(aes(x= Latitude_DD, y = Longitude_DD)  ) +
  geom_point(aes(size = Cyclopoida, shape = Month,
                 color = River), alpha = 0.2) +
  geom_line(aes(color = River), alpha = 0.4) +
  theme(legend.position = "none") +
  labs(x= "Latitude", y= "Longitude") +
  theme_bw() -> cyclop

all_data %>% 
  ggplot(aes(x= Latitude_DD, y = Longitude_DD)  ) +
  geom_point(aes(size = Chironomidae, shape = Month,
                 color = River), alpha = 0.2) +
  geom_line(aes(color = River), alpha = 0.4) +
  theme(legend.position = "none") +
  labs(x= "Latitude", y= "Longitude") +
  theme_bw() -> chiro


all_data %>% 
  ggplot(aes(x= Latitude_DD, y = Longitude_DD)  ) +
  geom_point(aes(size = Harpacticoida, shape = Month,
                 color = River), alpha = 0.2) +
  geom_line(aes(color = River), alpha = 0.4) +
  theme(legend.position = "none") +
  labs(x= "Latitude", y= "Longitude") +
  theme_bw() -> harp

cyclop 
chiro 
harp
```

For more species:
```{r}
all_data %>% 
  select(1:26, "Latitude_DD", "Longitude_DD", "River", "Month") %>% 
# remove the Unknown and some that appear only at one site
  select(!c("Unkown", "Coleoptera")) %>% 
  
  pivot_longer(cols = - c("Month", "River", "Latitude_DD", "Longitude_DD"),
               values_to = "values", names_to = "species") %>% 
  
# make it relative to the total abundance of this species  
  group_by(species) %>% 
  mutate(rel_abund = values/sum(values)) %>% 
  ungroup() %>% 
  
  ggplot(aes(x= Latitude_DD, y = Longitude_DD)  ) +
  geom_point(aes(size = rel_abund, shape = Month,
                 color = River), alpha = 0.2) +
  geom_line(aes(color = River), alpha = 0.4) +
   theme_bw() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right") +
  facet_wrap(~species) +
  labs(x= "Latitude", y= "Longitude") 
 
```
Or dont make it relative, use total numbers:

```{r}
all_data %>% 
  select(1:26, "Latitude_DD", "Longitude_DD", "River", "Month") %>% 
# remove the Unknown and some that appear only at one site
  select(!c("Unkown", "Coleoptera")) %>% 
  
  pivot_longer(cols = - c("Month", "River", "Latitude_DD", "Longitude_DD"),
               values_to = "values", names_to = "species") %>% 

  
  ggplot(aes(x= Latitude_DD, y = Longitude_DD)  ) +
  geom_point(aes(size = values, shape = Month,
                 color = River), alpha = 0.2) +
  geom_line(aes(color = River), alpha = 0.4) +
   theme_bw() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right") +
  facet_wrap(~species) +
  labs(x= "Latitude", y= "Longitude") 
```

The first species is most abundant in May in the Kardhiq rievr, and in September in Kardhiq and Vjosa.
The second species is most abundant in Drinos and Bence river. The third species is abundant in Disnica.